---
title: "RMS Titanic"
author: "Leslie Bostwick"
date: 2025-2-5
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- (List all variables here)
- (Class, Sex, Age, Survived, n)

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
- Are there any differences?
  - Yes there are differences in the data set (2201 people) versus wiki's (2224). 
- If yes, what might account for those differences?
  - These differences could be caused by stowways or crew not counted properly or lost records. 
  - ...

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(Class,n, fill= Sex))+geom_col()


```

**Observations**:

- Crew had the highest number of people survived at over 200, followed by 1st class (~200), 3rd (~175),  & 2nd (~120). A lot more females than males survived in the passenger categories versus crew. 

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(Class,Prop, fill= Sex))+geom_col()


```

**Observations**:

- Write your observations here.
- A larger percent of 1st class survived than any other category. This was followed by 2nd, 3rd, & crew. 
- A larger perent of females survived versus males.
- Is there anything *fishy* going on in your plot?
- Over 300% of 1st class survived which isn't possible
 

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}

df_prop %>% 
  filter(Survived == "Yes") %>%
  filter(Sex == "Female") %>% 
  ggplot(aes(Class,Prop, fill = Sex))+geom_col() + facet_grid(rows = "Age")
```

```{r q5-task2}
df_prop %>% 
  filter(Survived == "Yes") %>%
  filter(Sex == "Male") %>% 
  ggplot(aes(Class,Prop, fill = Sex))+geom_col() + facet_grid(rows = "Age")

```

**Observations**:

- All the children in 1st & 2nd class survived despite gender
- Almost all women survived across classes
- A low percent of men survived across classes
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - The graph in q4 added the adult category and child category together with them also separated by gender too. 

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.

**Research Observation**
- It's very easy to skew data to fit a certain narrative
- Think about what narrative the author of a graph is trying to portray and the assumptions that come with that narrative. 



