---
title: "Massachusetts Highway Stops"
author: "Leslie Bostwick"
date: 2025-04-27
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup
<!-- -------------------------------------------------- -->

### __q1__ Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "./data/yg821jf8611_ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA
<!-- -------------------------------------------------- -->

### __q2__ Do your "first checks" on the dataset. What are the basic facts about this dataset?
```{r q2}
head(df_data)
summary(df_data)

```

**Observations**:

- What are the basic facts about this dataset?
- During the data collection of this dataset
  - In the state of MA, the number of MA vehicles pull over was 3053713, then CT with 82906, NY with 69059, NH with 51514, and then RI with 39375. There is also 109857 in the other category and 9814 NA. 
  - There were 55830 searches conducted
  - There were 3602 frisk performed
  - This data was collected from 1-1-2007 to 12-31-2015
  - Twice as many men got stopped versus female
  - White people were stopped the most, followed by Black people, Hispanic people, and then Asian/Pacific islander. However 11008 citation were listed as other and there were 916 NAs.
  - There were no pedestrian stops which makes sense as this was a data set on highway stops
  - The youngest age pulled over was 10, the median age was 34, the mean age was 36.47 with the oldest age that was pulled over was 94. However there are 158006 NAs in this column, so a good portion of this pull over are missing this information.
  - In this data set, there are a lot of NAs through which is concern as there is lost information in that missing info. Whether it was a officer being lazy, was unsure or trying to hide something, we do not know nor will ever. 
- ...

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

- `race_Raw` could be the race of the police officer in the stop
- `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### __q3__ Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
level_subject <-
df_data %>% 
  summarise(unique(df_data$subject_race))

level_raw <-
df_data %>% 
  summarise(unique(df_data$raw_Race))


level_subject
level_raw

```

**Observations**:

-   What are the unique values for `subject_race`?
    -   asian/pacific islander
    -   black
    -   hispanic
    -   white
    -   other
    -   unknown
-   What are the unique values for `raw_Race`?
    -   A
    -   American Indian or Alaskan Native
    -   Asian or Pacific Islander
    -   Black
    -   Hispanic
    -   Middle Eastern or East Indian (South Asian)
    -   None - for no operator present citations only
    -   White
-   What is the overlap between the two sets?
    -   Black
    -   Hispanic
    -   White
    -   Asian or Pacific Islanders
-   What is the difference between the two sets?
    -   raw_race uses terms for admin purposes and non descriptiveness abbreviations like "A"
    -   raw_race includes specific labels while subject_race is more standardized with a consistent format

### __q4__ Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.

df_data <-
  df_data %>% 
  mutate(
    raw_Race_rename = case_when(
      raw_Race %in% c("White") ~ "white",
      raw_Race %in% c("Black") ~ "black",
      raw_Race %in% c("Hispanic") ~ "hispanic",
      raw_Race %in% c("Asian or Pacific Islander") ~ "asian/pacific islander",
      raw_Race %in% c("American Indian or Alaskan Native", 
                      "Middle Eastern or East Indian (South Asian)") ~ "other",
      raw_Race %in% c("None - for no operator present citations only", "A") ~ "unknown",
      TRUE ~ NA_character_
    )
  )

race_match_result <-
  df_data %>% 
  mutate(race_match = subject_race == raw_Race_rename) %>% 
  count(race_match) %>% 
  mutate(percentage = n / sum(n) * 100)

race_match_result
```

**Observations**

Between the two hypotheses:

- `race_Raw` could be the race of the police officer in the stop
- `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

- I think race_Raw is a unprocessed version of Subject_race as they have a 98.06% match to one another

## Vis
<!-- ------------------------- -->

### __q5__ Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)
```{r q5 1}
df_q5 <-
  df_data %>% 
  mutate(arrest_made = ifelse(is.na(arrest_made), FALSE, arrest_made))

df_q5 %>% 
  filter(!is.na(subject_age)) %>% 
  group_by(subject_age) %>% 
  summarize(arrest_rate = mean(arrest_made)) %>% 
  ggplot(aes(x = subject_age, y = arrest_rate)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
```

```{r q5 2}
df_q5 %>% 
  filter(!is.na(subject_sex)) %>% 
  group_by(subject_sex) %>% 
  summarise(arrest_rate = mean(arrest_made)) %>% 
  ggplot(aes(x = subject_sex, y = arrest_rate, fill = subject_sex)) +
  geom_col() 
```

```{r q5 3}
df_q5 %>% 
  filter(!is.na(subject_race)) %>% 
  group_by(subject_race) %>% 
  summarise(arrest_rate = mean(arrest_made)) %>% 
  ggplot(aes(x = subject_race, y = arrest_rate, fill = subject_race)) +
  geom_col()
```

**Observations**:

- How does `arrest_rate` tend to vary with `subject_age`?
  - A lot of 15 year olds are getting arrested at a higher rate than any other age. This might be because you have to be 16 to get a learner's permit in MA. The peak of arrest_rate is at 15 years and generally, the older you are, the lower the arrest rate is.
- How does `arrest_rate` tend to vary with `subject_sex`?
  - Gerernally, the arrest rate is higher for males than females
- How does `arrest_rate` tend to vary with `subject_race`?
  - Hispanic have the highest arrest rate, then black, then other, then white and finally asian/pacific Islander

# Modeling
<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### __q6__ Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

- Which `subject_race` levels are included in fitting the model?
  - The model includes white, black, and Hispanic individuals
- Which `subject_race` levels have terms in the model?
  - The terms only list out whites and hispanics, suggesting that black individuals are being used as the reference. 

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### __q7__ Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
 
fit_q7 <-
df_data %>% 
filter(
    !is.na(arrest_made),
    subject_race %in% c("white", "black", "hispanic")
) %>%   
mutate(
    subject_race = fct_relevel(subject_race, "white")
) %>% 
glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    family = "binomial"
  )

fit_q7 %>% tidy()
```

**Observations**:

- Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
  - Hispanic individuals are the most likely highest probabilty followed by black individual and then with the lowest probability white individuals.
- What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
  - Geography
  - Policing bias
  - The political climate at the of being pulled over
- Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
  - There is a contraband_found variables and recreational weed was not legal back when this data set was being collected unlike present day in MA so police could have been seeking it out. 

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### __q8__ Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8 <-
df_data %>% 
filter(
    !is.na(arrest_made),
    subject_race %in% c("white", "black", "hispanic")
) %>%   
mutate(
    subject_race = fct_relevel(subject_race, "white"),
) %>% 
glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
    family = "binomial"
  )

fit_q8 %>% tidy()
```

**Observations**:

- How does controlling for found contraband affect the `subject_race` terms in the model?
  - For the estimate compare to White individuals, Hispanics individuals dropped to .22 while Black individuals drop to -.05 so they were even less likely to have contraband found that white individuals.  
- What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
  - It tells us that a search occurred that was likely to lead to an arrest.
  - It does not tell us why the police decided to conduct a search. 

### __q9__ Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

Are racial disparities in arrest driven significantly by if searchs are conducted rather than difference in actual behavior or contraband possession?

```{r q9}
fit_q9 <-
  df_data %>% 
    filter(
     !is.na(search_conducted),
     subject_race %in% c("white","black","hispanic")
    ) %>% 
   mutate(
     subject_race = fct_relevel(subject_race, "white"),
     search_conducted = ifelse(is.na(search_conducted), FALSE, search_conducted)
   ) %>% 
   glm(
     formula = search_conducted ~ subject_age + subject_race + subject_sex,
     family = "binomial"
    )

fit_q9 %>% 
  tidy()


```

**Observations**:

- Hispanic individuals are more likely to get searched versus their black counterparts
- Female are less likely to get searched
- Younger people are slight more likely to be search than older people
- Personal police bias against hispanics could be a reason why they're more likely to get search especially in today's political climate.

## Further Reading
<!-- -------------------------------------------------- -->

- Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
