---
title: "Cookie Taste using Cohen's Kappa"
author: "Leslie, Oliver, Maya, Maya, Sparsh"
date: 2025-04-21
output:
  github_document:
    toc: true
---
## Set Up
```{r setup}
library(tidyverse)
library(modelr)
library(broom)
library(irr)
```

## Background Research
- What’s your question?
- What data do you plan to use to answer this question?
- What challenges do you anticipate encountering?
- What level of complexity of final product are you aiming for?
- What figures / tables do you anticipate producing?


## **Import file**

```{r import-task}
filename_cookie_taste <- "./data/Taste_Data.csv"
## NOTE: No need to edit
df_cookie_taste <-
  read_csv(
    filename_cookie_taste,
  )

df_cookie_taste
```

## **Conduct Initial Checks - EDA**

```{r intial-checks}
summary(df_cookie_taste)

head(df_cookie_taste)

```

#### Quantify sources of uncertainty
- How reputable is the source? How well-documented are their data-collection procedures? What potential errors might be in the data? What factors could contribute to missingness in the data?
- How were the data collected? Is it a random sample? A non-random sample? A full population?
- To the extent possible, report all estimates with confidence / prediction / tolerance intervals. If not possible, clearly explain why it is not possible to provide intervals and document what sources of uncertainty are not quantified.

#### Obeservations
- ....
- ....

## **Cohen's Kappa**

How to interception Cohen's Kappa from Geeks for Geeks

Kappa:

k=1: Perfect agreement beyond chance.
k=0: Agreement equal to that expected by chance alone.
k=−1: Perfect disagreement beyond chance.

####Interpretation of Cohen's Kappa values
Almost Perfect Agreement (0.81 - 1.00):
- Indicates very high agreement between the raters.
- Almost all observed agreement is due to actual agreement, with minimal disagreement.

Substantial Agreement (0.61 - 0.80):
- Represents a strong level of agreement between raters.
- A significant portion of the observed agreement is beyond what would be expected by chance.

Moderate Agreement (0.41 - 0.60):
- Suggests a moderate level of agreement.
- There is agreement, but there is still a notable amount of variability that cannot be attributed to agreement alone.

Fair Agreement (0.21 - 0.40):
- Indicates a fair level of agreement.
- Some agreement is present, but it may not be strong, and a substantial amount of variability exists.

Slight Agreement (0.00 - 0.20):
- Represents a slight level of agreement.
- The observed agreement is minimal, and most of it could be due to chance.

Poor Agreement (< 0.00):
- Signifies poor agreement, meaning the observed agreement is less than what would be expected by chance alone.


```{r Kappa}
df_taste <- df_cookie_taste %>% 
  select(-Name)

df_taste2 <- t(df_taste)

kappam.light(df_taste2)

```

```{r Yumminess}
df_yummy <-
df_taste %>% 
select(starts_with("Yum"))

df_yummy <- t(df_yummy)

kappam.light(df_yummy)
```
```{r Flavor}
df_flavor <-
df_taste %>% 
select(starts_with("Flavor"))

df_flavor <- t(df_flavor)

kappam.light(df_flavor)
```

```{r Texture}
df_texture <-
df_taste %>% 
select(starts_with("Text"))

df_texture <- t(df_texture)

kappam.light(df_texture)
```

```{r Chocolatey}
df_chocolatey <-
df_taste %>% 
select(starts_with("Chocolatey"))

df_chocolatey <- t(df_chocolatey)

kappam.light(df_chocolatey)
```

```{r Crispiness}
df_crispiness <-
df_taste %>% 
select(starts_with("Crispiness"))

df_crispiness <- t(df_crispiness)

kappam.light(df_crispiness)
```

```{r Chewiness}
df_chewiness <-
df_taste %>% 
select(starts_with("Chewiness"))

df_chewiness <- t(df_chewiness)

kappam.light(df_chewiness)
```
### Maybe the core team has more agreement
```{r Core}
df_core <-
df_cookie_taste %>% 
  filter(Name == "Oliver" | Name == "Crane" | Name == "Maya" | Name == "Sparsh") %>% 
  select(-Name)

df_core2 <- t(df_core)

kappam.light(df_core2)
```

```{r yummy2}
df_yummy2 <-
df_core %>% 
select(starts_with("Yum"))

df_yummy2 <- t(df_yummy2)

kappam.light(df_yummy2)

```

## **Graph Time**
```{r graph1}
df_graph <-
df_cookie_taste %>% 
select(Name,starts_with("Yum")) %>% 
pivot_longer(
    names_to = "Yummy",
    values_to = "n",
    cols = c('Yumminess/Satisfaction-5CNF', 'Yumminess/Satisfaction-10CNF', 'Yumminess/Satisfaction-15CNF', 'Yumminess/Satisfaction-5CF', 'Yumminess/Satisfaction-10CF', 'Yumminess/Satisfaction-15CF', 'Yumminess/Satisfaction-5EF', 'Yumminess/Satisfaction-10EF', 'Yumminess/Satisfaction-15EF', 'Yumminess/Satisfaction-5ENF','Yumminess/Satisfaction-10ENF', 'Yumminess/Satisfaction-15ENF')
  ) %>% 
mutate(
    Yummy = fct_reorder(Yummy, n)
  ) 

df_graph %>% 
ggplot(aes(Yummy, n, color = Name)) +
  geom_point(size = 2 , position = position_jitter(w = 0.2, h = 0)) +
  labs(
    x = "Yumminess",
    y = "Rating",
    color = "Raters"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r graph2}
df_graph %>% 
ggplot(aes(Yummy, n)) +
  geom_boxplot() +
  labs(
    x = "Yumminess",
    y = "Rating"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```