---
title: "Cookie Quality using Gage R&R"
author: "Leslie, Oliver, Maya, Maya, Sparsh"
date: 2025-04-21
output:
  github_document:
    toc: true
---
## Set Up
```{r setup}
library(tidyverse)
library(modelr)
library(broom)
```

## Background Research
- What’s your question?
- What data do you plan to use to answer this question?
- What challenges do you anticipate encountering?
- What level of complexity of final product are you aiming for?
- What figures / tables do you anticipate producing?


## **Import file**

```{r import-task}
filename_cookie <- "./data/results_gage_R_R.csv"
## NOTE: No need to edit
df_cookie <-
  read_csv(
    filename_cookie,
  )

df_cookie
```

## **Conduct Initial Checks - EDA**

```{r intial-checks}
summary(df_cookie)

head(df_cookie)
```

#### Quantify sources of uncertainty
- How reputable is the source? How well-documented are their data-collection procedures? What potential errors might be in the data? What factors could contribute to missingness in the data?
- How were the data collected? Is it a random sample? A non-random sample? A full population?
- To the extent possible, report all estimates with confidence / prediction / tolerance intervals. If not possible, clearly explain why it is not possible to provide intervals and document what sources of uncertainty are not quantified.
#### Obeservations
- ....
- ....


## **Conduct ANOVA**

### Circumference

```{r anova-circumference}
cookie.cir <- aov(Circumference ~ Index + Tray + Fridge + Cheap + Chocolate_chips + Data_taker, data = df_cookie)

summary(cookie.cir)
```

```{r homoscedasticity-circumference}
par(mfrow=c(2,2))
plot(cookie.cir)
par(mfrow=c(1,1))
```
### Height
```{r anova-height}
cookie.hei <- aov(Height ~ Index + Tray + Fridge + Cheap + Chocolate_chips + Data_taker, data = df_cookie)

summary(cookie.hei)
```

```{r homoscedasticity-height}
par(mfrow=c(2,2))
plot(cookie.hei)
par(mfrow=c(1,1))
```

### Length

```{r anova-length}
cookie.len <- aov(Length ~ Index + Tray + Fridge + Cheap + Chocolate_chips + Data_taker, data = df_cookie)

summary(cookie.len)
```

```{r homoscedasticity-length}
par(mfrow=c(2,2))
plot(cookie.len)
par(mfrow=c(1,1))
```

### Width

```{r anova-width}
cookie.wid <- aov(Width ~ Index + Tray + Fridge + Cheap + Chocolate_chips + Data_taker, data = df_cookie)

summary(cookie.wid)
```

```{r homoscedasticity-width}
par(mfrow=c(2,2))
plot(cookie.wid)
par(mfrow=c(1,1))
```

### Weight

```{r anova-weight}
cookie.wei <- aov(Weight ~ Index + Tray + Fridge + Cheap + Chocolate_chips + Data_taker, data = df_cookie)

summary(cookie.wei)
```

```{r homoscedasticity-weight}
par(mfrow=c(2,2))
plot(cookie.wei)
par(mfrow=c(1,1))
```
## **Tukey Honestly Significant Difference**
```{r Tukey HSD1}
tukey.cir<-TukeyHSD(cookie.cir)
tukey.hei<-TukeyHSD(cookie.hei)
tukey.len<-TukeyHSD(cookie.len)
tukey.wid<-TukeyHSD(cookie.wid)
tukey.wei<-TukeyHSD(cookie.wei)

tukey.cir
tukey.hei
tukey.len
tukey.wid
tukey.wei
```

## **Mean of Outputs**
```{r mean}
mean.cir <- df_cookie %>%
  group_by(Index, Cheap) %>%
  summarise(
      cir = mean(Circumference)
  )

mean.hei <- df_cookie %>%
  group_by(Index, Chocolate_chips) %>%
  summarise(
      hei = mean(Height)
  )

mean.len <- df_cookie %>%
  group_by(Index, Cheap) %>%
  summarise(
      len = mean(Length)
  )

mean.wid <- df_cookie %>%
  group_by(Index, Cheap) %>%
  summarise(
      wid = mean(Width)
  )

mean.wei <- df_cookie %>%
  group_by(Index,Cheap) %>%
  summarise(
      wei = mean(Weight)
  )

mean.cir
mean.hei
mean.len
mean.wid
mean.wei
```
## **Graph Time**

```{r graph1}
df_cookie %>%
ggplot(aes(as.factor(Index), Circumference, color = as.factor(Cheap))) +
  geom_point(cex = 1.5, pch = 1.0,position = position_jitter(w = 0.1, h = 0)) +
  stat_summary(fun.data = 'mean_se', geom = 'errorbar', width = 0.4) +
  stat_summary(fun.data = 'mean_se', geom = 'pointrange') +
  labs(
    x = "Cookie Index Number",
    y = "Circunfermce (cm)",
    color = "Cheap or Expensive"
  ) +
  theme_minimal()
```
```{r graph2}
df_cookie %>%
ggplot(aes(as.factor(Index), Height, color = as.factor(Chocolate_chips))) +
  geom_point(cex = 1.5, pch = 1.0,position = position_jitter(w = 0.1, h = 0)) +
  stat_summary(fun.data = 'mean_se', geom = 'errorbar', width = 0.4) +
  stat_summary(fun.data = 'mean_se', geom = 'pointrange') +
  labs(
    x = "Cookie Index Number",
    y = "Height (mm)",
    color = "Number of Chocolate chips"
  ) +
  theme_minimal()
```

```{r graph3}
df_cookie %>% 
ggplot(aes(as.factor(Index), Length, color = as.factor(Cheap))) +
  geom_point(cex = 1.5, pch = 1.0,position = position_jitter(w = 0.1, h = 0)) +
  stat_summary(fun.data = 'mean_se', geom = 'errorbar', width = 0.4) +
  stat_summary(fun.data = 'mean_se', geom = 'pointrange') +
  labs(
    x = "Cookie Index Number",
    y = "Length (mm)",
    color = "Cheap or Expensive"
  ) +
  theme_minimal()
```

```{r graph4}
df_cookie %>% 
ggplot(aes(as.factor(Index), Width, color = as.factor(Cheap))) +
  geom_point(cex = 1.5, pch = 1.0,position = position_jitter(w = 0.1, h = 0)) +
  stat_summary(fun.data = 'mean_se', geom = 'errorbar', width = 0.4) +
  stat_summary(fun.data = 'mean_se', geom = 'pointrange') +
  labs(
    x = "Cookie Index Number",
    y = "Width (mm)",
    color = "Cheap or Expensive"
  ) +
  theme_minimal()
```

```{r graph5}
df_cookie %>% 
ggplot(aes(as.factor(Index), Weight, color = as.factor(Cheap))) +
  geom_point(cex = 1.5, pch = 1.0,position = position_jitter(w = 0.1, h = 0)) +
  stat_summary(fun.data = 'mean_se', geom = 'errorbar', width = 0.4) +
  stat_summary(fun.data = 'mean_se', geom = 'pointrange') + 
  labs(
    x = "Cookie Index Number",
    y = "Weight after baking (grams)",
    color = "Cheap or Expensive"
  ) +
  theme_minimal()
```
